---
title: "Transit data viz - Santiago Subway called Metro"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, error=FALSE}
library(ggplot2)
library(osmdata)
library(sf)
library(ggspatial)
library(rio)
library(here)
library(dplyr)
library(ggthemes)
```

# Define directory 

```{r}
# Define GTFS directory 
# routes <- read.csv("data/routes.txt")
gtfs_dir <- here("gtfs_metro")

gtfs_data <- list.files(gtfs_dir, pattern = "\\.txt$", full.names = TRUE) %>%
  setNames(gsub("\\.txt$", "", basename(.))) %>%
  lapply(import)

list2env(gtfs_data, envir = .GlobalEnv)

lapply(gtfs_data, head)
```

# Find Metro's Stops & Routes, Metro's Shapes & Routes 

## Metro stops & routes

The GTFS dataset contains information for both bus, metro/subway, LRT, commuter rail or/and other public transport. In this analysis, we need obtain the stops and routes which belong to Metro. The Metro stops and routes are characterized as follows:
a) Metro Stops (stops_metro): Metro stops have purely alphabetical or special character IDs. Hence, the stop_id does not contain any digits (!grepl("\\d", stop_id)). 
b) Metro Routes (routes_metro): Metro routes are identified by a pattern like "L1", "L2", "L10", etc. Therefore, the route_id starts with "L" followed by a digit (grepl("^L\\d", route_id)).

```{r}
stops_metro <- stops %>% filter(!grepl("\\d", stop_id))
routes_metro <- routes %>% filter(grepl("^L\\d", route_id))
```

## Metro shapes & routes 

```{r}
# The goal here is to select Metro shape entries that correspond to Metro routes.
shapes_metro <- shapes %>% 
  filter(shape_id %in% trips$shape_id[trips$route_id %in% routes_metro$route_id]) %>% # it shows the relations amongst txt file 
  arrange(shape_id, shape_pt_sequence)
```

# Get colour for each Metro lines

First, get colour for whole network 

```{r}
# Combine Shape - trip - route 
shapes_colors <- left_join(left_join(shapes %>% select(shape_id) %>% unique(),
                                     trips %>% select(shape_id, route_id) %>% unique(),
                                     by = "shape_id"),
                           routes %>% select(route_id, route_color) %>% unique(),
                           by = "route_id") %>%
  mutate(route_color = paste0("#", route_color))

# Step 1: Extract unique shape IDs from shapes dataset
shapes_unique <- shapes %>% select(shape_id) %>% unique()

# Step 2: Extract unique shape-to-route mappings from trips dataset
trips_unique <- trips %>% select(shape_id, route_id) %>% unique()

# Step 3: Join shapes with trip data to get route_id
shapes_with_routes <- left_join(shapes_unique, trips_unique, by = "shape_id")

# Step 4: Extract unique route_id and route_color from routes dataset
routes_unique <- routes %>% select(route_id, route_color) %>% unique()

# Step 5: Join the shapes_with_routes data with route colors
shapes_colors <- left_join(shapes_with_routes, routes_unique, by = "route_id")

# Step 6: Format route_color to ensure it has a '#' prefix
shapes_colors <- shapes_colors %>% mutate(route_color = paste0("#", route_color))
```

Then, get colour for metro only. The code creates `shapes_colors_metro`, a filtered version of `shapes_colors` that includes only shape IDs associated with Metro routes. It does this by first identifying `shape_id`s that belong to trips where the `route_id` is in `routes_metro`, ensuring that only Metro-related shapes are retained. The `unique()` function removes any duplicate rows that may exist after the filtering process. Finally, the dataset is arranged by `shape_id` to maintain an organized structure, making it easier to analyze and visualize Metro route colors systematically.

```{r}
shapes_colors_metro <- shapes_colors %>%
  filter(shape_id %in% trips$shape_id[trips$route_id %in% routes_metro$route_id]) %>% unique() %>%
  arrange(shape_id)
```

# Plot 

```{r}
p1 <- ggplot() +
  geom_path(data=shapes, aes(shape_pt_lon, shape_pt_lat, group=shape_id), color="white", size=.2, alpha=.05) +
  geom_path(data=shapes_metro, aes(shape_pt_lon, shape_pt_lat, group=shape_id, colour=shape_id), size = 2, alpha=.7) +
  scale_color_manual(values=shapes_colors_metro$route_color) +
  geom_point(data=stops_metro, aes(stop_lon, stop_lat), shape=21, colour="white", alpha =.8) +
  coord_equal() +
  theme_void() +  # Replacing theme_null() with theme_void()
  theme(plot.background = element_rect(fill = "black", colour = "black"),
        title = element_text(hjust=1, colour="white", size = 8),
        axis.title.x = element_text(hjust=0, colour="white", size = 7)) +
  # xlab(sprintf("Joshua Kunst | Jkunst.com %s",format(Sys.Date(), "%Y"))) +
  ggtitle("TRANSANTIAGO\nSantiago's public transport system") +
  theme(legend.position = "none") 
p1
```

```{r}
ggsave(filename = "plot/gtfs-transantiago.pdf", plot = p1, bg = "black")
```
