---
title: "gtfstools"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Reading and manipulating GTFS files with data.table 

```{r}
library(gtfstools)
path <- system.file("extdata/spo_gtfs.zip", package = "gtfstools")
gtfs <- read_gtfs(path) # transform to object, data.table 
names(gtfs)
head(gtfs$trips)
```

```{r}
# Using data.table to manipulate the dataset 
original_headway <- gtfs$frequencies$headway_secs # save original data
head(gtfs$frequencies, 3)

gtfs$frequencies[, headway_secs := headway_secs + 100] # update headway + 100s
# gtfs$frequencies <- gtfs$frequencies$headway_secs + 100 
head(gtfs$frequencies, 3)

# restores the original headway
gtfs$frequencies[, headway_secs := original_headway]
head(gtfs$frequencies, 3)
```

```{r}
# write modified GTFS objects to GTFS txt files again
export_path <- tempfile("new_gtfs", fileext = ".zip")  # points to the path where gtfs should be written to 
write_gtfs(gtfs, path = export_path) # write gtfs to path 
zip::zip_list(export_path)[, c("filename", "compressed_size", "timestamp")] # list files within feed 
```

### Calculating trip speed

```{r}
# calculating the speeds of all trips (km/h)
speeds <- get_trip_speed(gtfs)
head(speeds)
nrow(speeds)
```

```{r}
# calculates the speeds of two specific trips (km/h)
speeds <- get_trip_speed(gtfs, trip_id = c("CPTM L07-0", "2002-10-0"))
speeds
```

```{r}
# calculate the length of all trips (km)
lengths <- get_trip_length(gtfs, file = "shapes")
head(lengths)
```

```{r}
# calculates the duration of all trips (minute)
durations <- get_trip_duration(gtfs)
head(durations)
```

### Combining and filtering feeds

#### Combine / merge GTFS objects that have same data table from 2 gtfs

```{r}
# Remove duplication of record
gtfs$agency
gtfs$calendar

no_dups_gtfs <- remove_duplicates(gtfs)
no_dups_gtfs$agency
no_dups_gtfs$calendar
```

```{r}
# merge 2 feeds file: bus feed and rail feed for one aree
poa_path <- system.file("extdata/poa_gtfs.zip", package = "gtfstools") # reads Porto Alegre's GTFS
poa_gtfs <- read_gtfs(poa_path)
poa_gtfs$agency

# combines Porto Alegre's and São Paulo's GTFS objects
combined_gtfs <- merge_gtfs(no_dups_gtfs, poa_gtfs)

# check result 
combined_gtfs$agency
```

#### Filtering by identifiers

```{r}
# checks pre-filter object size 
utils::object.size(gtfs)
```

```{r}
# select column with data.table 
head(gtfs$trips[, .(trip_id, trip_headsign, shape_id)])

# keeps entries related to the two specified ids
filtered_gtfs <- filter_by_trip_id(
  gtfs,
  trip_id = c("CPTM L07-0", "CPTM L07-1")
)

# checks post-filter object size
utils::object.size(filtered_gtfs)

# check data after filter
head(filtered_gtfs$trips[, .(trip_id, trip_headsign, shape_id)])
unique(filtered_gtfs)
unique(filtered_gtfs$shapes$shape_id)
```
 
#### Filtering by day of the week and time of the day

```{r}
# keeps services that operate on both saturday AND sunday
head(no_dups_gtfs$calendar)

filtered_gtfs <- filter_by_weekday(
  no_dups_gtfs,
  weekday = c("saturday", "sunday"),
  combine = "and"
)

filtered_gtfs$calendar[, c("service_id", "sunday", "saturday")]

# keeps services that operate on both saturday OR sunday
filtered_gtfs <- filter_by_weekday(
  no_dups_gtfs,
  weekday = c("saturday", "sunday"),
  combine = "or"
)

filtered_gtfs$calendar[, c("service_id", "sunday", "saturday")]
```


```{r}
# filter by time of the day => apply for "frequencies"
filtered_gtfs <- filter_by_time_of_day(gtfs, from = "05:00:00", to = "06:00:00")
head(filtered_gtfs$frequencies)
```

#### Filtering using a spatial extent

```{r}
# Ex: SPTran's Feed, shape = 68962
library(ggplot2)
# creates a polygon with the bounding box of shape 68962
# note: shape of GTFS presents a path that vehicle follows. 

shape_68962 <- convert_shapes_to_sf(gtfs, shape_id = "68962") # convert shape to a simple feature spatial linestring object representing the route's geometry.
bbox <- sf::st_bbox(shape_68962) # bbox is now a numeric vector containing the min/max coordinates that frame the shape (xmin: Minimum longitude (left), ymin: Minimum latitude (bottom), xmax: Maximum longitude (right), ymax: Maximum latitude (top))
bbox_geometry <- sf::st_as_sfc(bbox) # a polygon. While st_bbox() gives a basic numerical extent, converting it into an sfc polygon makes it more flexible for spatial analysis

# creates a geometry with all the shapes described in the gtfs
all_shapes <- convert_shapes_to_sf(gtfs)

ggplot() +
  geom_sf(data = all_shapes) +
  geom_sf(data = bbox_geometry, fill = NA, color = "red") +
  theme_minimal()
```

```{r}
# filter_by_sf() keeps all entries related to trips that INTERSECT with the specified polygon bbox (bbox was calculated above shape = 68962)
filtered_gtfs <- filter_by_sf(gtfs, bbox) # This function filters the GTFS dataset (gtfs) to include only elements intersect the bounding box (bbox).

# An intersection occurs when:
## A shape (route line) passes through the bbox (even partially).
## A shape starts or ends inside the bbox.
## A shape completely fits inside the bbox.
## A shape only touches the edge of the bbox.

filtered_shapes <- convert_shapes_to_sf(filtered_gtfs)

ggplot() +
  geom_sf(data = filtered_shapes) +
  geom_sf(data = bbox_geometry, fill = NA, color = "red") +
  theme_minimal()
```

```{r}
# try with spatial_operation = sf::st_contains 
filtered_gtfs <- filter_by_sf(gtfs, bbox, spatial_operation = sf::st_contains)
filtered_shapes <- convert_shapes_to_sf(filtered_gtfs)

ggplot() +
  geom_sf(data = filtered_shapes) +
  geom_sf(data = bbox_geometry, fill = NA, color = "red") +
  theme_minimal()
```

### Validating GTFS data

```{r}
tmpdir <- tempdir()

validator_path <- download_validator(tmpdir)
validator_path
```

```{r, eval=FALSE}
output_dir <- tempfile("gtfs_validation")

validate_gtfs(
  path,
  output_path = output_dir,
  validator_path = validator_path
)

list.files(output_dir)
```

### workflow example: spatial visualization of headways

1. How are the times between vehicles operating the same route (the headways) spatially distributed in SPTrans’ GTFS?
2. Or Time differences between consecutive trips serving the same stop along the same route

```{r}
# Scope of analysis: services operating during the morning peak, between 7am and 9am, on a typical tuesday
gtfs <- read_gtfs(path)

library(dplyr)
# filters the GTFS
filtered_gtfs <- gtfs |>
  remove_duplicates() |> 
  filter_by_weekday("tuesday") |> 
  filter_by_time_of_day(from = "07:00:00", to = "09:00:00")

# checking the result
filtered_gtfs$frequencies[trip_id == "2105-10-0"] # check time of the date
filtered_gtfs$calendar
```

Data shows that each trip is associated to more than one headway, as shown above (one entry for the 7am to 7:59am interval and another for the 8am to 8:59am interval). To solve this, average headway from 7am to 9am is calculated. 

Calculate the average headways weighted by the time duration of each headway:
- Step 1: multiply each headway by the size of the time interval during which it is valid. To calculate the time intervals,first use the convert_time_to_seconds() function to calculate the start and end time of the time interval in seconds and then subtract the latter by the former.
- Step 2: sum these multiplication results for each trip
- Step 3: and then divide this amount by the total time interval (two hours, in our case)

```{r}
filtered_gtfs <- convert_time_to_seconds(filtered_gtfs)
filtered_gtfs$frequencies

# check how the results look like for a particular trip id
filtered_gtfs$frequencies[trip_id == "2105-10-0"]

filtered_gtfs$frequencies[, time_interval := end_time_secs - start_time_secs]

average_headway <- filtered_gtfs$frequencies[,
  .(average_headway = weighted.mean(x = headway_secs, w = time_interval)),
  by = trip_id
]

average_headway[trip_id == "2105-10-0"]

head(average_headway)
```

```{r, eval=FALSE}
# calculate step by step (alternative option)
library(data.table)

# Sample data
df <- data.table(
  trip_id = c("2105-10-0", "2105-10-0"),
  start_time = c("07:00:00", "08:00:00"),
  end_time = c("07:59:00", "08:59:00"),
  headway_secs = c(900, 1200)
)

# Function to convert time to seconds since midnight
convert_time_to_seconds <- function(time_str) {
  hms <- as.numeric(strsplit(time_str, ":")[[1]])
  return(hms[1] * 3600 + hms[2] * 60 + hms[3])
}

# Convert start and end times to seconds
df[, start_seconds := sapply(start_time, convert_time_to_seconds)]
df[, end_seconds := sapply(end_time, convert_time_to_seconds)]

# Calculate time interval duration
df[, time_interval := end_seconds - start_seconds]

# Compute weighted average headway
weighted_avg_headway <- df[, sum(headway_secs * time_interval) / sum(time_interval)]

# Print the result
print(weighted_avg_headway)
```

The importance of weighted mean is: If we used a simple mean, we’d assume both headways are equally relevant. But if one headway applies for a longer period, it should have more influence on the final result. This method gives a more accurate reflection of what passengers experience.

Next, visualisation of the headways (time between veicles)

```{r}
# generate each trip geometry and join this data to the average headways
selected_trips <- average_headway$trip_id  # char vector contain trip id 

geometries <- get_trip_geometry(
  filtered_gtfs,  # gtfs data contain files, e.g. trip, shape, calendar, frequencies, stops,...
  trip_id = selected_trips,
  file = "shapes"
)

head(geometries)

# Note: get_trip_geometry() function returns the spatial geometries of the trips in the feed. This function allows us to specify which trips we want to generate the geometries of, so we are only going to apply the procedure to the trips present in the average headways table:
```

```{r}
# join the average headway data to the geometries and then configure the map 
geoms_with_headways <- merge(
  geometries,
  average_headway,
  by = "trip_id"
)
```

```{r}
# the color and line width of each trip geometry varies with its headway
ggplot(geoms_with_headways) +
  geom_sf(aes(color = average_headway, size = average_headway), alpha = 0.8) +
  scale_color_gradient(high = "#132B43", low = "#56B1F7") +
  labs(color = "Average headway", size = "Average headway") +
  theme_minimal()
```
